---
title: "Bikesharing Dataset"
author: "Geethika Satheesh & Venkata Satya Sreeya Ch"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r Packages Installation}
suppressWarnings(
  suppressPackageStartupMessages({
library(dplyr)
library(lubridate)
library(tidyverse)
library(skimr)
library(janitor)
library(corrplot)
library (ggplot2)
library(forecast)
library(zoo)
library(TTR)
library(gridExtra)
library(scales)
library(tseries)
library(forecast)
library(prophet)
  })
)
```

## Load Data

```{r}
day_data <- read.csv("day.csv")
hour_data <- read.csv("hour.csv")

#shape of data
dim(day_data)
dim(hour_data)

#structure of data
glimpse(day_data)
glimpse(hour_data)

#summary of data
summary(day_data)
summary(hour_data)

```


```{r}
head(day_data)
head(hour_data)
tail(day_data)
tail(hour_data)
```

##Data Cleaning 

```{r}
# ── STEP 1: Check NA values (on raw data - perfectly fine here) ──
colSums(is.na(day_data))
colSums(is.na(hour_data))

# Check blank empty cells
sapply(day_data[, sapply(day_data, is.character)], 
       function(x) sum(x == "", na.rm = TRUE))
sapply(hour_data[, sapply(hour_data, is.character)], 
       function(x) sum(x == "", na.rm = TRUE))

# ── STEP 2: Convert date columns ──
day_data$dteday  <- as.Date(day_data$dteday,  format = "%m/%d/%Y")
hour_data$dteday <- as.Date(hour_data$dteday, format = "%m/%d/%Y")

#Full timestamp - completely unambiguous for R
hour_data$datetime <- as.POSIXct(
  sprintf("%s %02d:00:00", 
          as.character(hour_data$dteday), 
          as.integer(hour_data$hr)),
  format = "%Y-%m-%d %H:%M:%S"
)

# ── STEP 3: Verify ──
str(day_data$dteday)     # Should show: Date[1:731]
str(hour_data$datetime)  # Should show: POSIXct[1:17379]
head(hour_data$datetime) # Preview first few timestamps

all_days      <- seq(min(day_data$dteday), 
                     max(day_data$dteday), by = "day")
missing_days  <- setdiff(all_days, day_data$dteday)

all_hours     <- seq(min(hour_data$datetime), 
                     max(hour_data$datetime), by = "hour")
missing_hours <- setdiff(all_hours, hour_data$datetime)

cat("Missing days :", length(missing_days),  "\n")
cat("Missing hours:", length(missing_hours), "\n")

# ── Inspect Missing Hours ──
missing_df <- data.frame(
  hour  = as.integer(format(as.POSIXct(missing_hours, origin="1970-01-01"), "%H")),
  date  = as.Date(as.POSIXct(missing_hours, origin="1970-01-01")),
  month = as.integer(format(as.POSIXct(missing_hours, origin="1970-01-01"), "%m"))
)

print(table(missing_df$hour))   # Missing by hour
print(table(missing_df$month))  # Missing by month
print(sort(table(missing_df$date), decreasing=TRUE)[1:20]) # Top dates


#handling missing values
# Confirm: 2011-01-27 had 22/24 hours missing = full system shutdown
# Confirm: 2012-10-29 & 10-30 = Hurricane Sandy documented event

# DECISION: Use Linear Interpolation
# Reason: Missing hours are during extreme weather when
# rentals would naturally be near 0 anyway

full_grid <- data.frame(
  datetime = seq(min(hour_data$datetime),
                 max(hour_data$datetime),
                 by = "hour")
)

hour_complete <- merge(full_grid, hour_data,
                       by = "datetime", all.x = TRUE)

# Interpolate all key numeric columns
cols_to_fix <- c("cnt","casual","registered",
                 "temp","atemp","hum","windspeed")

hour_complete[cols_to_fix] <- lapply(
  hour_complete[cols_to_fix], 
  na.approx, na.rm = FALSE
)

hour_complete <- hour_complete[!is.na(hour_complete$hr), ]
# Verify
cat("Rows after completion:", nrow(hour_complete), "\n") # 17544
cat("Remaining NAs in cnt:", sum(is.na(hour_complete$cnt)), "\n") # 0
```

##Outlier detection

```{r}
# Daily data - key numeric columns
par(mfrow = c(2, 3))
boxplot(day_data$cnt,       main = "Total Rentals (cnt)",  col = "steelblue")
boxplot(day_data$casual,    main = "Casual Users",          col = "tomato")
boxplot(day_data$registered,main = "Registered Users",      col = "seagreen")
boxplot(day_data$temp,      main = "Temperature",           col = "orange")
boxplot(day_data$hum,       main = "Humidity",              col = "purple")
boxplot(day_data$windspeed, main = "Wind Speed",            col = "pink")

#STEP 9: IQR-based outlier count
detect_outliers <- function(x) {
  Q1  <- quantile(x, 0.25, na.rm = TRUE)
  Q3  <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  sum(x < (Q1 - 1.5 * IQR) | x > (Q3 + 1.5 * IQR), na.rm = TRUE)
}

# Apply to all numeric columns in day_data
num_cols <- sapply(day_data, is.numeric)
outlier_counts <- sapply(day_data[, num_cols], detect_outliers)
print(outlier_counts)

```


```{r}
# Likely sensor errors (humidity near 0 is physically impossible)
day_data$hum[day_data$hum < 0.1] <- median(day_data$hum)
# Verify 
boxplot(day_data$hum, main = "Humidity After Fix", col = "purple")
min(day_data$hum)  # Should no longer be near 0

# ── Create cleaned copy ──
day_clean <- day_data

# ── Define event dates ──
event_dates <- as.Date(c(
  "2012-10-29",  # Hurricane Sandy Day 1
  "2012-10-30",  # Hurricane Sandy Day 2
  "2011-01-27",  # Winter Storm (22/24 hours missing)
  "2011-01-18",  # Winter Storm
  "2011-08-28"   # Hurricane Irene
))

# ── Replace event cnt values with NA then interpolate ──
day_clean$cnt[day_clean$dteday %in% event_dates]        <- NA
day_clean$casual[day_clean$dteday %in% event_dates]     <- NA
day_clean$registered[day_clean$dteday %in% event_dates] <- NA

day_clean$cnt        <- na.approx(day_clean$cnt,        na.rm = FALSE)
day_clean$casual     <- na.approx(day_clean$casual,     na.rm = FALSE)
day_clean$registered <- na.approx(day_clean$registered, na.rm = FALSE)

# ── Verify ──
cat("Original Hurricane Sandy cnt:", 
    day_data$cnt[day_data$dteday == "2012-10-29"], "\n")
cat("Cleaned  Hurricane Sandy cnt:", 
    day_clean$cnt[day_clean$dteday == "2012-10-29"], "\n")
```



```{r}
#Exploratory Data analysis

# Trend & Seasonality (Time Plot with Trend)
ggplot(day_clean, aes(x = dteday, y = cnt)) +
  geom_line(color = "grey70") +
  geom_smooth(method = "loess",formula=y~x, color = "red", 
              se = TRUE, linewidth = 1.2) +
  labs(title    = "Daily Bike Demand (2011-2012)",
       subtitle = "Upward trend + seasonal waves observed",
       x = "Date", y = "Total Rentals") +
  theme_minimal()
```


"The time plot reveals a strong upward trend in daily bike rentals from early 2011 through mid 2012, followed by a seasonal decline. Two complete annual seasonal cycles are observed with seasonal amplitude increasing proportionally with the trend level, indicating a multiplicative rather than additive pattern. Significant day-to-day volatility suggests weather and calendar effects drive short-term fluctuations, justifying the inclusion of exogenous variables in SARIMAX and Prophet models."



```{r}
#Monthly + Weekly Demand Side by Side

# Monthly
p1 <- day_clean %>%
  group_by(mnth) %>%
  summarise(avg = mean(cnt)) %>%
  ggplot(aes(x = factor(mnth), y = avg, fill = factor(mnth))) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_x_discrete(labels = month.abb) +
  labs(title = "Monthly Demand Pattern",
       x = "Month", y = "Avg Rentals") +
  theme_minimal()

# Weekly
p2 <- day_clean %>%
  group_by(weekday) %>%
  summarise(avg = mean(cnt)) %>%
  ggplot(aes(x = factor(weekday), y = avg, 
             fill = factor(weekday))) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_x_discrete(labels = c("Sun","Mon","Tue",
                               "Wed","Thu","Fri","Sat")) +
  labs(title = "Weekly Demand Pattern",
       x = "Day", y = "Avg Rentals") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2,
  top = "Supply Allocation: Monthly & Weekly Patterns")
```

"Monthly analysis reveals strong seasonal demand variation with peak rentals in June–September (~5500–5800) and minimum demand in January–February (~2200), directly informing seasonal supply allocation. Weekly patterns show relatively uniform demand across days, suggesting that day-of-week alone is insufficient for supply decisions — hourly analysis being more critical for operational planning."


```{r}
#Intraday + Weekday vs Weekend Hourly

hour_complete$day_type <- ifelse(
  hour_complete$weekday %in% c(0,6), "Weekend", "Weekday"
)

hour_complete %>%
  group_by(hr, day_type) %>%
  summarise(avg_cnt = mean(cnt, na.rm = TRUE), 
            .groups = "drop") %>%
  ggplot(aes(x = hr, y = avg_cnt, 
             color = day_type, group = day_type)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 1.8) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_manual(values = c("Weekday" = "steelblue",
                                 "Weekend" = "tomato")) +
  labs(title    = "Intraday Demand: Weekday vs Weekend",
       subtitle = "Weekday = commute peaks | Weekend = leisure midday peak",
       x = "Hour of Day", y = "Average Rentals",
       color = "") +
  theme_minimal()
```

"Intraday analysis reveals fundamentally different demand patterns between weekdays and weekends. Weekdays exhibit a bimodal distribution with sharp peaks at 8AM and 5–6PM corresponding to morning and evening commutes, while weekends show a unimodal leisurely peak between 12–3PM. This distinction directly informs supply redistribution strategies — commuter hubs require stocking before 8AM and 4PM on weekdays, while leisure zones need peak supply around midday on weekends. The 3–4AM window represents the optimal maintenance period across both day types."


```{r}
#Hour x Weekday Heatmap
hour_complete %>%
  mutate(day_name = factor(weekday, levels = 0:6,
                            labels = c("Sun","Mon","Tue",
                                       "Wed","Thu","Fri","Sat"))) %>%
  group_by(hr, day_name) %>%
  summarise(avg_cnt = mean(cnt, na.rm = TRUE), 
            .groups = "drop") %>%
  ggplot(aes(x = day_name, y = hr, fill = avg_cnt)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low      = "white",
                        mid      = "steelblue",
                        high     = "darkred",
                        midpoint = 200) +
  scale_y_continuous(breaks = 0:23) +
  labs(title    = "Demand Heatmap: Hour × Day of Week",
       subtitle = "Dark red = critical supply zones for waiting time reduction",
       x = "Day of Week", y = "Hour",
       fill = "Avg Rentals") +
  theme_minimal()
```

"The hour × weekday demand heatmap precisely identifies critical supply zones for waiting time reduction. The darkest red zones at hours 8 and 17–18 on weekdays represent morning and evening commute peaks exceeding 500 average rentals, confirming these as highest waiting time risk periods. Wednesday and Thursday evenings represent the absolute peak demand moments in the dataset. Weekend demand is uniformly distributed across midday hours without sharp peaks. Early morning hours 0–6 consistently show near-zero demand across all days, representing the optimal window for bike maintenance and station rebalancing operations."



```{r}
#Weather Impact (Temp + Weather Situation)
p1 <- ggplot(day_clean, aes(x = temp * 41, y = cnt)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  labs(title = "Temperature vs Demand",
       x = "Temperature (°C)", y = "Rentals") +
  theme_minimal()

p2 <- ggplot(day_clean, aes(x = factor(weathersit), y = cnt,
                              fill = factor(weathersit))) +
  geom_boxplot(show.legend = FALSE) +
  scale_x_discrete(labels = c("Clear","Misty","Rain/Snow")) +
  scale_fill_manual(values = c("#59a14f","#f28e2b","#e15759")) +
  labs(title = "Weather Condition vs Demand",
       x = "Weather", y = "Rentals") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2,
  top = "Q: Does Weather Drive Demand?")
```

"Weather conditions significantly influence bike sharing demand. Temperature exhibits a non-linear inverted-U relationship with demand, peaking around 25°C before declining at higher temperatures, necessitating its inclusion as an exogenous variable in SARIMAX and Prophet models. Weather situation analysis reveals that rain and snow conditions reduce median daily demand by approximately 60% compared to clear weather conditions, while misty conditions cause a moderate 15% reduction. These findings strongly justify the use of weather variables as external regressors in advanced forecasting models."


```{r}
#Casual vs Registered User Behavior
day_clean %>%
  mutate(day_name = factor(weekday, levels = 0:6,
                            labels = c("Sun","Mon","Tue",
                                       "Wed","Thu","Fri","Sat"))) %>%
  group_by(day_name) %>%
  summarise(Casual     = mean(casual),
            Registered = mean(registered)) %>%
  tidyr::pivot_longer(cols      = c(Casual, Registered),
                      names_to  = "type",
                      values_to = "avg") %>%
  ggplot(aes(x = day_name, y = avg, fill = type)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Casual"     = "tomato",
                                "Registered" = "steelblue")) +
  labs(title    = "Casual vs Registered Users by Day",
       subtitle = "Registered = commuters (weekdays) | Casual = leisure (weekends)",
       x = "Day", y = "Avg Rentals", fill = "") +
  theme_minimal()
```

"Registered users consistently dominate demand across all days of the week, comprising approximately 75–85% of total rentals on weekdays, confirming their role as regular commuters. Casual users exhibit a distinct weekend surge, nearly doubling from weekday levels (~600) to weekend levels (~1400–1500), indicating leisure and tourist ridership. This behavioral divergence necessitates differentiated supply strategies — weekday supply should prioritize commuter corridors and transit hubs, while weekend supply should target recreational and tourist areas to minimize waiting time for both user segments."


```{r}
#demand surge vs surplus over time (gap between actual and 7-day rolling average)
day_clean$rolling_supply <- rollmean(day_clean$cnt,
                                      k     = 7,
                                      fill  = NA,
                                      align = "right")
day_clean$supply_gap <- day_clean$cnt - day_clean$rolling_supply
day_clean %>%
  filter(!is.na(supply_gap)) %>%
  ggplot(aes(x = dteday)) +
  geom_bar(aes(y    = supply_gap,
               fill = supply_gap > 0),
           stat = "identity") +
  scale_fill_manual(
    values = c("TRUE"  = "darkred",
               "FALSE" = "steelblue"),
    labels = c("TRUE"  = "Demand Surge (Waiting Risk)",
               "FALSE" = "Demand Drop (Surplus)")) +
  
  # ✅ Meaningful date breaks
  scale_x_date(
    breaks       = seq(as.Date("2011-01-01"),
                       as.Date("2012-12-31"),
                       by = "2 months"),
    date_labels  = "%b\n%Y",        # Jan\n2011 format
    minor_breaks = "1 month") +
  
  # ✅ Meaningful y axis
  scale_y_continuous(
    breaks = seq(-5000, 2000, by = 1000),
    labels = comma) +
  
  # ✅ Add reference line at 0
  geom_hline(yintercept = 0, 
             color      = "black", 
             linewidth  = 0.8) +
  
  # ✅ Annotate key events
  annotate("text", x = as.Date("2012-10-29"), y = -4200,
           label = "Hurricane\nSandy", 
           color = "black", size = 3, fontface = "bold") +
  annotate("text", x = as.Date("2011-01-27"), y = -2800,
           label = "Winter\nStorm",
           color = "black", size = 3, fontface = "bold") +
  
  labs(title    = "Demand Surge vs Surplus Over Time",
       subtitle = "Red = Waiting Time Risk | Blue = Surplus Opportunity",
       x        = "Date",
       y        = "Gap (Actual − 7-Day Rolling Average)",
       fill     = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 0.5))
```


Red bars consistently ~+500 to +2000 throughout → persistent waiting time risk
Blue drops much deeper (-3000 to -4500) → extreme weather creates massive surplus
Hurricane Sandy (Oct 2012) → deepest blue drop clearly visible ~-4500
Winter Storm (Jan 2011) → second deepest ~-2800
Pattern is symmetric — surges and drops alternate regularly throughout both years


```{r}
# Create timeseries  object
# Create week_data
week_data <- day_clean %>%
  mutate(week = format(dteday, "%Y-%U")) %>%
  group_by(week) %>%
  summarise(cnt = sum(cnt), .groups = "drop")

# Create month_data 
month_data <- day_clean %>%
  group_by(yr, mnth) %>%
  summarise(cnt = sum(cnt), .groups = "drop")
ts_daily   <- ts(day_clean$cnt,  frequency = 365, start = c(2011,1))
ts_weekly  <- ts(week_data$cnt,  frequency = 52,  start = c(2011,1))
ts_monthly <- ts(month_data$cnt, frequency = 12,  start = c(2011,1))

# Decompose & Plot
# Daily STL
stl_daily <- stl(ts_daily, 
                  s.window = "periodic", 
                  robust   = TRUE)
plot(stl_daily, main = "STL — Daily")

# Weekly STL
stl_weekly <- stl(ts_weekly, 
                   s.window = "periodic", 
                   robust   = TRUE)
plot(stl_weekly, main = "STL — Weekly")

# Monthly Classical only
# STL fails — only 2 periods (24 months)
plot(decompose(ts_monthly, type = "multiplicative"))
```

STL Daily
Data: Full observed series with high volatility ✅
Seasonal: Weekly cycles clearly visible, amplitude changes over time → confirms STL was right choice over classical
Trend: Perfectly smooth linear growth from ~2500 → 6500 ✅
Remainder: Small random noise throughout, one large spike at end 2012 → Hurricane Sandy residual effect still visible

STL Weekly
Data: Much smoother than daily ✅
Seasonal: Annual wave pattern clearly visible
Trend: Clean linear upward growth ~18000 → 42000 weekly rentals ✅
Remainder: Mostly flat with few isolated spikes → extreme weather events

Monthly Multiplicative
Observed: Clear inverted U shape — rises then falls ✅
Trend: Smooth growth up to mid 2012 then decline
Seasonal: Ratio fluctuates 0.7 → 1.15 → confirms multiplicative pattern
Random: Flat line with edge data loss → classical decomposition limitation


```{r}

adf_daily   <- adf.test(ts_daily)
adf_weekly  <- adf.test(ts_weekly)
adf_monthly <- adf.test(ts_monthly)

cat("Daily   p-value:", round(adf_daily$p.value,   4), "\n")
cat("Weekly  p-value:", round(adf_weekly$p.value,  4), "\n")
cat("Monthly p-value:", round(adf_monthly$p.value, 4), "\n")

```

All p-values > 0.05 → All three need differencing before ACF & PACF


```{r}
#Apply First Order Differencing
# ── First Difference ──
diff_daily   <- diff(ts_daily,   differences = 1)
diff_weekly  <- diff(ts_weekly,  differences = 1)
diff_monthly <- diff(ts_monthly, differences = 1)

# ── Re-test After Differencing ──
adf_diff_daily   <- adf.test(diff_daily)
adf_diff_weekly  <- adf.test(diff_weekly)
adf_diff_monthly <- adf.test(diff_monthly)

cat("After Differencing:\n")
cat("Daily   p-value:", round(adf_diff_daily$p.value,   4), "\n")
cat("Weekly  p-value:", round(adf_diff_weekly$p.value,  4), "\n")
cat("Monthly p-value:", round(adf_diff_monthly$p.value, 4), "\n")

```

weekly and monthly need second differencing, daily is now stationary

```{r}
# Second difference for weekly & monthly
diff2_weekly  <- diff(diff_weekly,  differences = 1)
diff2_monthly <- diff(diff_monthly, differences = 1)

# Re-test
adf2_weekly  <- adf.test(diff2_weekly)
adf2_monthly <- adf.test(diff2_monthly)

cat("Weekly  2nd diff p-value:", round(adf2_weekly$p.value,  4), "\n")
cat("Monthly 2nd diff p-value:", round(adf2_monthly$p.value, 4), "\n")
```

monthly needs second differencing, weekly is now stationary

```{r}
# Third difference for monthly
diff3_monthly <- diff(diff2_monthly, differences = 1)

adf3_monthly <- adf.test(diff3_monthly)
cat("Monthly 3rd diff p-value:", round(adf3_monthly$p.value, 4), "\n")
```

"Monthly data with only 24 observations was differenced once (d=1) as further differencing reduced the series to insufficient observations for reliable ADF testing, with the borderline p-value of 0.29 accepted given the small sample size limitation."


```{r}
par(mfrow = c(3,2), mar = c(3, 3, 3, 1))
acf(diff_daily,    lag.max = 30, main = "ACF — Daily (d=1)")
pacf(diff_daily,   lag.max = 30, main = "PACF — Daily (d=1)")
acf(diff2_weekly,  lag.max = 20, main = "ACF — Weekly (d=2)")
pacf(diff2_weekly, lag.max = 20, main = "PACF — Weekly (d=2)")
acf(diff_monthly,  lag.max = 12, main = "ACF — Monthly (d=1)")
pacf(diff_monthly, lag.max = 12, main = "PACF — Monthly (d=1)")
par(mfrow = c(1,1))

```


ACF and PACF analysis on differenced series revealed that daily and monthly series follow an ARIMA(1,1,1) pattern with autocorrelation cutting off after lag 1 in both plots. Weekly series after second differencing suggests an ARIMA(1,2,1) structure. These manually identified orders will be compared against auto.arima selections for final model validation."


```{r}
# TRAIN/TEST SPLIT
day_clean  <- day_clean[order(day_clean$dteday), ]
n          <- nrow(day_clean)
train_data <- day_clean[1:(n-60), ]
test_data  <- day_clean[(n-59):n, ]

cat("========== Split Summary ==========\n")
cat("Total days :", n, "\n")
cat("Train days :", nrow(train_data), "\n")
cat("Test  days :", nrow(test_data),  "\n")
cat("Train start:", as.character(min(train_data$dteday)), "\n")
cat("Train end  :", as.character(max(train_data$dteday)), "\n")
cat("Test  start:", as.character(min(test_data$dteday)),  "\n")
cat("Test  end  :", as.character(max(test_data$dteday)),  "\n")
cat("Overlap    :",
    length(intersect(train_data$dteday,
                     test_data$dteday)),
    "← must be 0\n")


#CREATE TS OBJECTS
# Holt-Winters (classroom)
ts_train_hw <- ts(train_data$cnt,
                   frequency = 7,
                   start     = c(1, 1))
ts_test_hw  <- ts(test_data$cnt,
                   frequency = 7,
                   start     = c(97, 1))

# Main models
ts_train_msts <- msts(train_data$cnt,
                       seasonal.periods = c(7, 365),
                       start            = c(2011, 1))
ts_test_msts  <- msts(test_data$cnt,
                       seasonal.periods = c(7, 365),
                       start            = c(2012, 306))

cat("HW    train:", length(ts_train_hw),   "\n")
cat("HW    test :", length(ts_test_hw),    "\n")
cat("MSTS  train:", length(ts_train_msts), "\n")
cat("MSTS  test :", length(ts_test_msts),  "\n")
```


```{r}
#CLASSROOM — HOLT-WINTERS

hw_model    <- HoltWinters(ts_train_hw,
                            seasonal = "multiplicative")
hw_forecast <- forecast(hw_model, h = 60)

# ── Plot ──
autoplot(hw_forecast) +
  autolayer(ts_test_hw,
            series    = "Actual",
            color     = "red",
            linewidth = 0.8) +
  labs(title    = "Holt-Winters MultiplicativeZ",
       subtitle = "Weekly seasonality (frequency=7)",
       x = "Time", y = "Total Rentals") +
  theme_minimal()

# ── Parameters ──
cat("=== Smoothing Parameters ===\n")
cat("Alpha:", round(hw_model$alpha, 4), "\n")
cat("Beta :", round(hw_model$beta,  4), "\n")
cat("Gamma:", round(hw_model$gamma, 4), "\n")

# ── Evaluate ──
hw_pred   <- as.numeric(hw_forecast$mean)
hw_actual <- as.numeric(ts_test_hw)

hw_mae  <- mean(abs(hw_actual - hw_pred))
hw_rmse <- sqrt(mean((hw_actual - hw_pred)^2))
hw_mape <- mean(abs((hw_actual - hw_pred)
                     / hw_actual)) * 100

cat("\n=== Holt-Winters Results ===\n")
cat("MAE :", round(hw_mae,  2), "\n")
cat("RMSE:", round(hw_rmse, 2), "\n")
cat("MAPE:", round(hw_mape, 2), "%\n")

```



```{r}

#SEASONAL NAIVE

snaive_model <- snaive(ts_train_msts, h = 60)
snaive_pred  <- as.numeric(snaive_model$mean)
actual       <- as.numeric(ts_test_msts)

# ── Plot ──
autoplot(snaive_model) +
  autolayer(ts_test_msts,
            series    = "Actual",
            color     = "red",
            linewidth = 0.8) +
  labs(title    = "Seasonal Naive Forecast",
       subtitle = "Multi-seasonal benchmark (7 & 365)",
       x = "Time", y = "Total Rentals") +
  theme_minimal()

# ── Evaluate ──
snaive_mae  <- mean(abs(actual - snaive_pred))
snaive_rmse <- sqrt(mean((actual - snaive_pred)^2))
snaive_mape <- mean(abs((actual - snaive_pred)
                         / actual)) * 100

cat("=== Seasonal Naive Results ===\n")
cat("MAE :", round(snaive_mae,  2), "\n")
cat("RMSE:", round(snaive_rmse, 2), "\n")
cat("MAPE:", round(snaive_mape, 2), "%\n")
```


```{r}

#TBATS

tbats_model    <- tbats(ts_train_msts)
tbats_forecast <- forecast(tbats_model, h = 60)
tbats_pred     <- as.numeric(tbats_forecast$mean)

# ── Model Summary ──
cat("TBATS Model:\n")
print(tbats_model)

# ── Plot ──
autoplot(tbats_forecast) +
  autolayer(ts_test_msts,
            series    = "Actual",
            color     = "red",
            linewidth = 0.8) +
  labs(title    = "TBATS Forecast",
       subtitle = "Multi-seasonal: weekly + annual",
       x = "Time", y = "Total Rentals") +
  theme_minimal()

# ── Evaluate ──
tbats_mae  <- mean(abs(actual - tbats_pred))
tbats_rmse <- sqrt(mean((actual - tbats_pred)^2))
tbats_mape <- mean(abs((actual - tbats_pred)
                        / actual)) * 100

cat("\n=== TBATS Results ===\n")
cat("MAE :", round(tbats_mae,  2), "\n")
cat("RMSE:", round(tbats_rmse, 2), "\n")
cat("MAPE:", round(tbats_mape, 2), "%\n")

# ── vs Seasonal Naive ──
cat("\nImprovement over Seasonal Naive:\n")
cat("MAPE:", round((snaive_mape - tbats_mape)/
                    snaive_mape * 100, 2), "%\n")
```


```{r}
# PHASE 9: ARIMA
arima_model <- auto.arima(ts_train_msts,
                           seasonal      = TRUE,
                           stepwise      = FALSE,
                           approximation = FALSE)

arima_forecast <- forecast(arima_model, h = 60)
arima_pred     <- as.numeric(arima_forecast$mean)

# ── Model Summary ──
cat("ARIMA Order Selected:\n")
print(summary(arima_model))

# ── Plot ──
autoplot(arima_forecast) +
  autolayer(ts_test_msts,
            series    = "Actual",
            color     = "red",
            linewidth = 0.8) +
  labs(title    = "ARIMA Forecast",
       subtitle = "Auto-selected optimal order",
       x = "Time", y = "Total Rentals") +
  theme_minimal()

# ── Evaluate ──
arima_mae  <- mean(abs(actual - arima_pred))
arima_rmse <- sqrt(mean((actual - arima_pred)^2))
arima_mape <- mean(abs((actual - arima_pred)
                        / actual)) * 100

cat("\n=== ARIMA Results ===\n")
cat("MAE :", round(arima_mae,  2), "\n")
cat("RMSE:", round(arima_rmse, 2), "\n")
cat("MAPE:", round(arima_mape, 2), "%\n")

cat("\nImprovement over Seasonal Naive:\n")
cat("MAPE:", round((snaive_mape - arima_mape)/
                    snaive_mape * 100, 2), "%\n")
```


```{r}
#PROPHET

library(prophet)

# ── Prepare data ──
prophet_train <- data.frame(
  ds          = train_data$dteday,
  y           = train_data$cnt,
  temp        = train_data$temp,
  hum         = train_data$hum,
  windspeed   = train_data$windspeed,
  weathersit  = train_data$weathersit
)

prophet_test <- data.frame(
  ds          = test_data$dteday,
  y           = test_data$cnt,
  temp        = test_data$temp,
  hum         = test_data$hum,
  windspeed   = test_data$windspeed,
  weathersit  = test_data$weathersit
)

# ── Build model ──
prophet_model <- prophet(
  seasonality.mode        = "multiplicative",
  changepoint.prior.scale = 0.05,
  yearly.seasonality      = TRUE,
  weekly.seasonality      = TRUE,
  daily.seasonality       = FALSE
)

# ── Add regressors ──
prophet_model <- add_regressor(prophet_model, "temp")
prophet_model <- add_regressor(prophet_model, "hum")
prophet_model <- add_regressor(prophet_model, "windspeed")
prophet_model <- add_regressor(prophet_model, "weathersit")

# ── Add US holidays ──
prophet_model <- add_country_holidays(prophet_model,
                                       country_name = "US")

# ── Fit ──
prophet_model <- fit.prophet(prophet_model,
                              prophet_train)

# ── Forecast ──
prophet_fc   <- predict(prophet_model, prophet_test)
prophet_pred <- prophet_fc$yhat

# ── Plot forecast ──
plot(prophet_model, prophet_fc) +
  labs(title = "Prophet Forecast — Multi-Seasonal + Weather + Holidays",
       x = "Date", y = "Total Rentals") +
  theme_minimal()

# ── Component plot ──
prophet_plot_components(prophet_model, prophet_fc)

# ── Evaluate ──
prophet_actual <- test_data$cnt
prophet_mae    <- mean(abs(prophet_actual - prophet_pred))
prophet_rmse   <- sqrt(mean((prophet_actual - prophet_pred)^2))
prophet_mape   <- mean(abs((prophet_actual - prophet_pred)
                            / prophet_actual)) * 100

cat("=== Prophet Results ===\n")
cat("MAE :", round(prophet_mae,  2), "\n")
cat("RMSE:", round(prophet_rmse, 2), "\n")
cat("MAPE:", round(prophet_mape, 2), "%\n")

cat("\nImprovement over Seasonal Naive:\n")
cat("MAPE:", round((snaive_mape - prophet_mape)/
                    snaive_mape * 100, 2), "%\n")
```


Prophet MAPE = 25.63%
Beats ALL baseline models



```{r}
# PHASE 11: XGBOOST

library(xgboost)

# ── Feature Engineering ──
create_features <- function(df) {
  df$month     <- as.numeric(format(df$dteday, "%m"))
  df$weekday_n <- as.numeric(df$weekday)
  df$lag1      <- dplyr::lag(df$cnt, 1)
  df$lag7      <- dplyr::lag(df$cnt, 7)
  df$lag30     <- dplyr::lag(df$cnt, 30)
  df$roll7     <- zoo::rollmean(df$cnt, 7,
                                 fill  = NA,
                                 align = "right")
  return(df)
}

# ── Apply to full data first ──
day_feat <- create_features(day_clean)
day_feat <- day_feat[!is.na(day_feat$lag30), ]

n_feat     <- nrow(day_feat)
train_feat <- day_feat[1:(n_feat-60), ]
test_feat  <- day_feat[(n_feat-59):n_feat, ]

# ── Features & Target ──
features <- c("temp","hum","windspeed",
               "weathersit","season",
               "month","weekday_n",
               "holiday","workingday",
               "lag1","lag7","lag30","roll7")

X_train <- as.matrix(train_feat[, features])
y_train <- train_feat$cnt
X_test  <- as.matrix(test_feat[, features])
y_test  <- test_feat$cnt

# ── DMatrix ──
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

# ── Parameters ──
xgb_params <- list(
  max_depth        = 4,
  eta              = 0.05,
  subsample        = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 5,
  gamma            = 0.1,
  lambda           = 1.0,
  objective        = "reg:squarederror"
)

# ── Cross Validation ──
xgb_cv <- xgb.cv(
  params                = xgb_params,
  data                  = dtrain,
  nrounds               = 500,
  nfold                 = 5,
  early_stopping_rounds = 20,
  verbose               = FALSE
)
best_rounds <- xgb_cv$best_iteration
if (length(best_rounds) == 0) {
  best_rounds <- 100  # Default fallback if CV fails
  message("Warning: CV did not return best_iteration. Using default of 100.")
}

cat("Best rounds:", best_rounds, "\n")

# ── Train ──
xgb_model <- xgb.train(
  params  = xgb_params,
  data    = dtrain,
  nrounds = best_rounds
)

# ── Predict ──
xgb_pred <- predict(xgb_model, dtest)

# ── Feature Importance ──
importance <- xgb.importance(feature_names = features,
                              model         = xgb_model)
xgb.plot.importance(importance,
                     main = "XGBoost Feature Importance")

# ── Overfitting Check ──
train_pred_xgb <- predict(xgb_model, dtrain)
cat("\nOverfitting Check:\n")
cat("Train RMSE:", round(sqrt(mean(
  (y_train - train_pred_xgb)^2)), 2), "\n")
cat("Test  RMSE:", round(sqrt(mean(
  (y_test  - xgb_pred)^2)), 2), "\n")

# ── Evaluate ──
xgb_mae  <- mean(abs(y_test - xgb_pred))
xgb_rmse <- sqrt(mean((y_test - xgb_pred)^2))
xgb_mape <- mean(abs((y_test - xgb_pred)
                       / y_test)) * 100

cat("\n=== XGBoost Results ===\n")
cat("MAE :", round(xgb_mae,  2), "\n")
cat("RMSE:", round(xgb_rmse, 2), "\n")
cat("MAPE:", round(xgb_mape, 2), "%\n")

cat("\nImprovement over Prophet:\n")
cat("MAPE:", round((prophet_mape - xgb_mape)/
                    prophet_mape * 100, 2), "%\n")
```

```{r}

#Model comparision
# ── Actual vs Predicted — All Models ──
comparison_df <- data.frame(
  Date           = test_data$dteday,
  Actual         = as.numeric(ts_test_msts),
  Seasonal_Naive = snaive_pred,
  TBATS          = tbats_pred,
  ARIMA          = arima_pred,
  Prophet        = prophet_pred,
  XGBoost        = xgb_pred
)

comparison_long <- comparison_df %>%
  pivot_longer(cols      = -Date,
               names_to  = "Model",
               values_to = "Value")

ggplot(comparison_long,
       aes(x     = Date,
           y     = Value,
           color = Model,
           group = Model)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(
    values = c("Actual"         = "black",
               "Seasonal_Naive" = "grey50",
               "TBATS"          = "orange",
               "ARIMA"          = "purple",
               "Prophet"        = "steelblue",
               "XGBoost"        = "darkred")) +
  scale_x_date(
    date_breaks  = "2 weeks",
    date_labels  = "%b %d") +
  labs(title    = "Actual vs Predicted — All Models",
       subtitle = "Test Period: Nov–Dec 2012 | Closer to Black = Better",
       x        = "Date",
       y        = "Total Rentals",
       color    = "Model") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x     = element_text(angle = 45,
                                        hjust = 1))
```


```{r}
# ── Final Metrics Table ──
results <- data.frame(
  Model = c("Holt-Winters",
            "Seasonal Naive",
            "TBATS",
            "ARIMA",
            "Prophet",
            "XGBoost"),
  Track = c("Classroom",
            "Main",
            "Main",
            "Main",
            "Main",
            "Main"),
  MAE   = round(c(hw_mae,
                   snaive_mae,
                   tbats_mae,
                   arima_mae,
                   prophet_mae,
                   xgb_mae), 2),
  RMSE  = round(c(hw_rmse,
                   snaive_rmse,
                   tbats_rmse,
                   arima_rmse,
                   prophet_rmse,
                   xgb_rmse), 2),
  MAPE  = paste0(round(c(hw_mape,
                          snaive_mape,
                          tbats_mape,
                          arima_mape,
                          prophet_mape,
                          xgb_mape), 2), "%")
)

# ── Order by RMSE ──
results <- results[order(results$RMSE), ]

# ── Print Table ──
cat("╔══════════════════════════════════════════════════╗\n")
cat("║           MODEL COMPARISON TABLE                ║\n")
cat("╠══════════════════════════════════════════════════╣\n")
cat(sprintf("%-18s %-10s %-10s %-10s %-10s\n",
            "Model", "Track", "MAE", "RMSE", "MAPE"))
cat("╠══════════════════════════════════════════════════╣\n")
for(i in 1:nrow(results)) {
  cat(sprintf("%-18s %-10s %-10s %-10s %-10s\n",
              results$Model[i],
              results$Track[i],
              results$MAE[i],
              results$RMSE[i],
              results$MAPE[i]))
}
cat("╠══════════════════════════════════════════════════╣\n")
cat(sprintf("Best Model → %s (RMSE: %s | MAPE: %s)\n",
            results$Model[1],
            results$RMSE[1],
            results$MAPE[1]))
cat("╚══════════════════════════════════════════════════╝\n")
```



```{r}
# ════════════════════════════════════
# PHASE 13: SUPPLY OPTIMIZATION
# ════════════════════════════════════

# ── Next 30 days forecast ──
future_dates <- data.frame(
  ds         = seq(as.Date("2013-01-01"), by="day", length.out=30),
  temp       = mean(train_data$temp),
  hum        = mean(train_data$hum),
  windspeed  = mean(train_data$windspeed),
  weathersit = 1
)

future_fc <- predict(prophet_model, future_dates)

# ── Build supply plan ──
supply_plan <- data.frame(
  Date     = future_dates$ds,
  Demand   = pmax(0, round(future_fc$yhat)),
  Lower    = pmax(0, round(future_fc$yhat_lower)),
  Upper    = pmax(0, round(future_fc$yhat_upper)),
  Supply   = pmax(0, round(future_fc$yhat * 1.15)),
  Risk     = ifelse(future_fc$yhat > quantile(future_fc$yhat, 0.75), "High",
             ifelse(future_fc$yhat > quantile(future_fc$yhat, 0.50), "Medium", "Low"))
)

# ── Print ──
cat("===== Daily Supply Plan — Jan 2013 =====\n")
print(supply_plan[, c("Date","Demand","Supply","Risk")], 
      row.names = FALSE)

cat("\n===== Monthly Summary =====\n")
cat("Total Demand :", sum(supply_plan$Demand),  "\n")
cat("Total Supply :", sum(supply_plan$Supply),  "\n")
cat("Buffer       :", sum(supply_plan$Supply) - 
                      sum(supply_plan$Demand),  "\n")
cat("Peak Day     :", max(supply_plan$Demand),  "\n")
cat("Lowest Day   :", min(supply_plan$Demand),  "\n")
cat("Avg Daily    :", round(mean(supply_plan$Demand)), "\n")

# ── Plot ──
ggplot(supply_plan, aes(x = Date)) +
  geom_ribbon(aes(ymin = Lower, ymax = Upper),
              fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = Demand, color = "Forecast"),
            linewidth = 1.2) +
  geom_line(aes(y = Supply, color = "Recommended Supply"),
            linewidth = 1.2, linetype = "dashed") +
  scale_color_manual(
    values = c("Forecast"            = "steelblue",
               "Recommended Supply"  = "darkred")) +
  scale_x_date(date_breaks = "1 week",
               date_labels = "%b %d") +
  labs(title    = "30-Day Bike Supply Plan — January 2013",
       subtitle = "Shaded = Uncertainty | Dashed = Recommended Supply (15% buffer)",
       x = "Date", y = "Bikes", color = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x     = element_text(angle = 45, hjust = 1))
```


```{r}
#RESIDUAL ANALYSIS
# ── Calculate residuals ──
residuals_prophet <- as.numeric(ts_test_msts) - prophet_pred
residuals_xgb     <- y_test - xgb_pred

par(mfrow = c(2,2))

# ── Prophet Residuals ──
plot(residuals_prophet, type = "l",
     main = "Prophet Residuals",
     col  = "steelblue",
     ylab = "Error", xlab = "Day")
abline(h = 0, col = "red", lwd = 1.5)

hist(residuals_prophet, breaks = 20,
     main = "Prophet Residual Distribution",
     col  = "steelblue", xlab = "Error")

# ── XGBoost Residuals ──
plot(residuals_xgb, type = "l",
     main = "XGBoost Residuals",
     col  = "darkred",
     ylab = "Error", xlab = "Day")
abline(h = 0, col = "red", lwd = 1.5)

hist(residuals_xgb, breaks = 20,
     main = "XGBoost Residual Distribution",
     col  = "darkred", xlab = "Error")

par(mfrow = c(1,1))
```


"Residual analysis confirms both Prophet and XGBoost produce errors centered near zero with no systematic bias. XGBoost residuals show a more symmetric distribution and random pattern, indicating better model fit. Both models exhibit a common large negative error around day 20 of the test period corresponding to pre-Thanksgiving demand surge in late November 2012, suggesting holiday transition effects are slightly underestimated. Overall residual patterns validate both models as reliable forecasting tools for bike sharing supply optimization."